---
title: "Basic Models"
author: "Rachael Caelie (Rocky) Aikens"
date: "6/10/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center")
source("../R/generate_data.R")
source("../R/diagnose_data.R")
library(tidyverse)
library(ggpubr)
library(knitr)
library(binom)
theme_set(theme_light())

n <- 50000
```

# A Simple Model with Diagnosis by Severity

Suppose we want to do a study of a disease. Perhaps our target population is "all people," or perhaps it is some subset of people (e.g. only Men, only Asians, etc.). Each person, $i$, has an underlying (unobserved) disease state $Z_i$ and a disease severity $S_i$.  A simple model is:

\begin{align*}
    Z_i &\sim_{iid} \text{Bernoulli}(0.15)\\
    S_i|Z_i = 1 &\sim_{iid} \text{Uniform}(0, 1)\\
\end{align*}

Where $S_i = 0$ whenever $Z_i = 0$. 

Suppose additionally that a person's likelihood of diagnosis is a function of the severity of their illness, namely

$$D_i \sim_{iid} \text{Bernoulli}\left(\phi(S_i\right))$$

$\phi(S_i)$ could be many possible functions, here I use a sigmoid function:

$$\phi(S_i) = \frac{1}{1 + exp(-(\beta_0 + \beta_1S_i))} - c$$
Here, $c = \frac{1}{1 + exp(-\beta_0)}$ is a corrective constant to ensure that $\phi(0) = 0$. Let $\beta_1 = 20$ and $\beta_0 = -10$.  This gives the following probability of diagnosis as a function of severity.

```{r plot diagnosis function, fig.height=3, fig.width=3}
S <- seq(0, 1, length.out = 100)

ggplot(data = data.frame(S), aes(x = S, y = diagnosis_fn(S, theta = c(-10, 20)))) +
  geom_line() + xlim(c(0,1)) + ylim(c(0,1)) + coord_fixed()+
  ylab("P(Diagnosis | S)")
```

A few example rows sampled from the above distribution are printed below:

```{r generate and diagnose data}
set.seed(123)

df <- generate_cross_sectional(n = n) %>%
  diagnose_cross_sectional(theta = c(-10, 20))

df %>% select(-x) %>% head()
```

\pagebreak

# Distortion from diagnostic process

This diagnosis process distorts the distribution of severities we observe.  

```{r plot diagnosed data}
diseased <- df %>% filter(disease == TRUE) %>%
  mutate(group = "Individuals with Disease", alpha = 0.7)

diagnosed <- df %>% filter(diagnosed == TRUE) %>%
  mutate(group = "Individuals Diagnosed", alpha = 1)

ggplot(rbind(diseased, diagnosed), aes(x = severity, group = group)) +
  geom_histogram(aes(alpha = alpha), position = "identity") +
  scale_alpha(range = c(0.4, 0.8))
```
